Economic Theory of Fertility Explains Modern Natural Selection

### 2020-12-01 why-negative-selection.Rmd
Add plot of mean PSEA by income; remove "causality" from appendix; nicer font

---
title: "Economic Theory of Fertility Explains Modern Natural Selection"
author: "David Hugh-Jones (D.Hugh-Jones@uea.ac.uk), Abdel Abdellaoui (a.abdellaoui@gmail.com)"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "Natural selection has been documented in contemporary humans, but little is known about the mechanisms behind it. We test for natural selection on 33 polygenic scores over two generations, using data from UK Biobank. Selection effects appear consistent over time. Polygenic scores which predict lower earnings and education are selected for. Effects are concentrated among low-income and less educated people, younger parents, people with more lifetime sexual partners, and people not living with a partner. Overall effect sizes are substantially larger after correcting for ascertainment bias. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns can be explained by economic theories of fertility, in which higher earnings may either increase or decrease fertility."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
---

@@

* Limit max no. of children in regressions (use an "in_children" bool)
  - not necessary. Very few have abnormally many.
* Check each result with a manually done regression!
* Consider doing separate regressions for "any children" vs. "n children | 1 or more"
  - heckit too slow and gives crazy answers, just use plm/ols
  - some variables do look different
* If we weight with the GHS, and plot the means of figure 1 using this w
  weighting, most things stay the same - but the change in EA is reversed! EA
  increases over time. Could this be because the sample includes many
  relatively educated older people? If so, then mean effects are *overestimated*
  by figure 1.
* Reweight other regressions (e.g. the omnibus regression in the appendix)?
* Look at geography, esp. in siblings regressions 
  - maybe Townshend f.189? Not sure essential
* Another point is that negative selection only among poor will increase
    genetic inequality.
* When we have better definition of "partner", bring it across from the other
  project to define with_partner
  - `with_partner` is fine, but if you want to have pairs who had the same children
  you need to do more
* If you take "year of last job started" and see if it is after/before
  the first child, you can figure out whether mothers keep working.
  - better still, look at all jobs started and see if any are within e.g. 5 years of
  first child's birth
  - this could then be another way to get at income vs substitution effects. You'd
  predict negative selection effects among mothers who worked. Of course the
  choice to work is endogenous.
* double check which scores are reverse coded!

@@

In addition, we do not know whether estimated effect sizes of natural
selection will increase or decrease as more accurate polygenic scores
are produced. We are also unsure whether genetic variants underlying other
phenotypes will show a similar pattern of natural selection as these ones. 
In short, it is probably too early to tell whether modern natural
selection is substantively important or not.

@@

## Causality

Different polygenic scores are correlated. Table
\@ref(tab:tbl-pgs-corrs) shows the top correlations in the sample.
Because of this, bivariate correlations between PGS and number of
children might be driven by other genetic scores. To explore which
polygenic scores are driving negative selection, we run a single omnibus
regression of *number of children* on all the PGS. We exclude EA2,
waist-hip ratio, waist-circumference, and "Hip combined" since they are
highly correlated with other scores, which could make our estimates
unstable. Figure \@ref(fig:plot-together) shows the results.
Interestingly, several PGS remain independently significant, although
effect sizes are reduced.

```{r tbl-pgs-corrs}
cor_tbl <- cor(famhist[score_names], use = "pair") %>% 
      as_tibble() %>% 
      mutate(score_name = score_names, .before = 1) %>% 
      pivot_longer(cols = -score_name, values_to = "Correlation") %>% 
      filter(score_name < name) %>% 
      arrange(desc(Correlation)) %>% 
      slice(1:10) %>% 
      mutate(across(c(score_name, name), pretty_names))

as_hux(cor_tbl, add_colnames = TRUE) %>% 
      set_contents(1, 1:2, c("PGS", "PGS")) %>% 
      theme_basic(header_col = FALSE) %>% 
      set_bottom_border(1, everywhere, 0.4) %>% 
      set_bottom_border(final(1), everywhere, 0.4) %>% 
      set_tb_padding(4) %>% 
      set_caption("Top 10 correlations between polygenic scores")

```{r plot-together, fig.cap = "Partial correlations of PGS with fertility, from a single regression."}

loadd(res_together)
n_regs <- as.double(nrow(res_together))
res_together %>% 
      mutate(
        term = pretty_names(term),
        dep.var = dplyr::recode(dep.var, 
          "n_children" = "N Children", "n_sibs" = "N Siblings"),
        dep.var = fct_relevel(dep.var, "N Siblings")
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(y = reorder_within(term, estimate, dep.var, order_abs(1)), 
        x = estimate, color = p < 0.05/{{n_regs}}, fill  = PCs)) + 
        pval_geom_point(alpha = 0.8) +
        my_vline +
        facet_wrap(vars(dep.var), scales = "free_y") +
        pval_scale() +
        my_fill_scale(n = 2) +
        my_labs() + 
        scale_y_reordered() +
        theme(legend.position = "bottom")
```

>>>>

---
title: "Economic Theory of Fertility Explains Modern Natural Selection"
author: "David Hugh-Jones (D.Hugh-Jones@uea.ac.uk), Abdel Abdellaoui (a.abdellaoui@gmail.com)"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "Natural selection has been documented in contemporary humans, but little is known about the mechanisms behind it. We test for natural selection on 33 polygenic scores over two generations, using data from UK Biobank. Selection effects appear consistent over time. Polygenic scores which predict lower earnings and education are selected for. Effects are concentrated among low-income and less educated people, younger parents, people with more lifetime sexual partners, and people not living with a partner. Overall effect sizes are substantially larger after correcting for ascertainment bias. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns can be explained by economic theories of fertility, in which higher earnings may either increase or decrease fertility."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Times
---

@@

* Limit max no. of children in regressions (use an "in_children" bool)
  - not necessary. Very few have abnormally many.
* Consider doing separate regressions for "any children" vs. "n children | 1 or more"
  - heckit too slow and gives crazy answers, just use plm/ols
  - some variables do look different
* If we weight with the GHS, and plot the means of figure 1 using this w
  weighting, most things stay the same - but the change in EA is reversed! EA
  increases over time. Could this be because the sample includes many
  relatively educated older people? If so, then mean effects are *overestimated*
  by figure 1.
* Reweight other regressions (e.g. the omnibus regression in the appendix)?
* Another point is that negative selection only among poor will increase
    genetic inequality.
  - one way to look at this: regress PGS on e.g. income or Quin71
  - then weight by n_children and see the change
  - needs to be comparable in some sense though, across generations
  - e.g. quintile of where born, unweighted; vs quintile of where now,
    weighted by children
  - perhaps simpler is just to run regressions with/without weights
    by n_children. This gives you the "effect" of natural selection
* If you take "year of last job started" and see if it is after/before
  the first child, you can figure out whether mothers keep working.
  - better still, look at all jobs started and see if any are within e.g. 5 years of
  first child's birth
  - this could then be another way to get at income vs substitution effects. You'd
  predict negative selection effects among mothers who worked. Of course the
  choice to work is endogenous.
* double check which scores are reverse coded!

@@

In addition, we do not know whether estimated effect sizes of natural
selection will increase or decrease as more accurate polygenic scores
are produced. We are also unsure whether genetic variants underlying other
phenotypes will show a similar pattern of natural selection as these ones. 
In short, it is probably too early to tell whether modern natural
selection has a substantively important effect on populations' mean 
polygenic scores.

Because selection effects are concentrated in poorer populations, they also increase
inequality with respect to polygenic scores. For example, Figure 
\@ref(fig:plot-mean-EA3-by-income) graphs children's mean polygenic scores for 
educational attainment (EA3), in households of different income groups. The blue
bars show the actual means. The grey bars show the hypothetical means if all 
households had equal numbers of children. Natural selection against EA3, which
lowers the mean, is stronger at the bottom of the income distribution, and this
increases the differences between groups.

```{r plot-mean-EA3-by-income, fig.align = "center", fig.cap = "Mean PGS for educational attainment (EA3) of children by household income group. Blue is actual. Grey is hypothetical in the absence of selection effects.", fig.width = 4}
income_EA3 <- famhist %>% 
      filter(! is.na(n_children), ! is.na(income_cat)) %>% 
      mutate(
        `Household income` = factor(income_cat, 
                          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", 
                                      "> £100K"))
      ) %>% 
      group_by(`Household income`) %>% 
      summarize(
        `Without selection` = mean(EA3_excl_23andMe_UK, na.rm = TRUE),
        Actual              = weighted.mean(EA3_excl_23andMe_UK, n_children, 
                                              na.rm = TRUE)
      ) %>% 
      tidyr::pivot_longer(-`Household income`) %>% 
      rename(
        `Children's mean EA3`     = value
      )

income_EA3 %>% 
              ggplot(aes(`Household income`, `Children's mean EA3`, fill = name)) +
                geom_col(position = "dodge") +
                scale_fill_manual(values = c("navy", "grey60")) +
                theme(
                  legend.background = element_rect(fill = "white", colour = NA),
                  legend.title = element_blank(), 
                  legend.position = c(.25, .85)
                )
```

@@
